Dish: Distribution-oblivious Shell Scripting

### 2019-11-28 pldi/draft.tex
Update graph components

% \begin{lstlisting}[language=sh,float=h,numbers=none]
% cat doc.ms |                   # print input file
% groff -t -e -mandoc -Tascii |  # remove formatting
% col -bx |                      # remove backspaces 
% tr A-Z a-z |                   # convert to lower case
% tr -d '[:punct:]' |            # remove punctuation
% sort |                         # put words in alphabetical order
% unique |                       # remove duplicate words
% comm -13 /usr/share/dict -     # report words not in dictionary 
% \end{lstlisting}
\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$[p_1]$$)
  sort | uniq -c | sort -rn | head 5 > out
\end{lstlisting}
\smallskip
% \noindent
Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
Combining several features, \unix makes small tasks easy to express;
  for a user with \emph{one} computer and small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
% composition allows the composition of general primitives

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
%\begin{wrapfigure}{r}{0.06\textwidth}
%  \vspace{-12pt}
%  % \begin{center}
%    % \includegraphics[width=0.48\textwidth]{gull}
%    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
%  % \end{center}
%  \vspace{-30pt}
%\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  ls -l | wc -l                          [($$p_2$$)]
\end{lstlisting}
\smallskip
Commands can be tuned by two mechanisms, namely com\-mand-line options and environment variables.
Options control the execution of a program---\eg changing the \ttt{sort}ing order.
The shell does not have any visibility into these options; 
  after it expands special characters such as \ttt{~} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.

@@

% At a high level, POSIX defines several file types:
% (-) regular files, storing lines of text or binary data,
% (d) directories, grouping together multiple other files,
% (l) symbolic links, pointing to other files,
% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  ls | sed 's/1/2/' | > out.t 2> err.t    [($$p_3$$)]
\end{lstlisting}
\smallskip

@@

\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  cat * | tee >(wc -l > l.t) | zip > a.z   [($$p_4$$)]
\end{lstlisting}
\smallskip

@@

\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  read | { calc | square; } >/dev/fd/0     [($$p_5$$)]
\end{lstlisting}
\smallskip

@@

\kk{Move some variation of that to intro/overview:

In the core of \sys there is a dataflow graph model that is inspired
by the ones that are used in popular distributed stream processing
systems (such as MapReduce \cite{mapreduce:08} and Apache Spark
\cite{spark:12}). These systems exploit data parallelism by
parallelizing nodes of the dataflow graph. Unfortunately, the
optimizations in many of these systems do not preserve the dataflow
model semantics, blurring the lines between specification,
optimization, and implementation. In contrast, our dataflow model
inherently supports the data parallelism that is found in shell
programs. More precisely, data parallelism can be easily exposed as
task parallelism in our model. We exploit that by developing a set of
parallelization-exposing optimizations that preserve the dataflow
model semantics, and can be represented as graph transformations. The
\sys optimizer then applies these transformations in an iterative way,
optimizing the program to exploit the available computational
resources.

@@

%% A very important consequence of this is that we can iteratively apply
%% optimizations that expose data parallelism, or even revert them when
%% the resulting dataflow graph is too wide. This flexibility unifies the
%% unoptimized and optimized versions of a computation, enabling thinking
%% about them with the same conceptual tools.
}

In order to effectively apply optimizations and distribute a shell
script, it first has to be translated to a manipulable representation.
In \Cref{graph-components} we propose a slight variation of the
standard dataflow graph model as a candidate intermediate
representation. The main benefit of our model is that it exposes both
task and data parallelism. In \Cref{ir:transformations} we define a
set of optimizations that expose and exploit data parallelism of
commands in \sta and \pur as semantics-preserving graph
transformations.

@@

As mentioned in \Cref{bg:pipelines}, the two main abstractions of the
shell are files that contain data, and commands that communicate
through these files. Based on that, we define nodes in our graph model
to represent commands, and edges to represent files.

\paragraph{Nodes - Commands}

For a set $D$, we write $D\kstar$ to denote the set of all finite
words over $D$. For words $x, y \in D\kstar$, we write $x \cdot y$
or $xy$ to denote their concatenation. We write $\eps$ for the empty
word. We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$. It is easy to see that $\leq$ is a partial order, and it is often referred to as the \emph{prefix order}.

@@

\noindent
This command takes a stream of lines from its standard input and
returns a stream of lines on its standard output. An important
observation is that commands in the shell might read their input
stream from a sequence of different sources. An illustrative example
is the command \ttt{cat} which reads its input from a combination
of files and its standard input.

@@

The example above reads file \ttt{x} until it encounters the EOF
character, then reads from \ttt{stdin}, and then reads from
\ttt{y}. In order to handle this particularity, nodes in the
dataflow graph can have many incoming edges, that are ordered in a
sequence. Note that this does not correspond to standard dataflow
graph representations where the semantics of multiple incoming edges
is either zipping the inputs in pairs, or arbitraliy interleaving them.

@@

Note that nodes can read from more files than just the files of their
stream (or respectively write to more files that their output stream),
however these files are static (e.g. used for configuration or
logging) and do not represent streams of data, and therefore are not
considered part of the dataflow graph. We assume that these static
files can only be accessed by one node, thus ensuring that they do not
interfere with the distributed implementation.
%% Because of the assumption in section Front End \ref{}, that each file
%% exists only once in the dataflow graph, we can safely assume that
%% reading and writing to these static files does not interfere with the
%% distributed implementation, and doesn't alter the behaviour of the
%% program.

\paragraph{Edges - Files}

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D\kstar$. They could either refer to a specific named file in the file
system, or just be FIFO pipes that used for interprocess communication
%% \TODO{Are there any other file types, such as URLs, ...?}.

@@

In order to expose data parallelism from the output of a command, files that
don't point to a resource can have
\begin{wrapfigure}[3]{r}{0.20\columnwidth}
\vspace{-7pt}
  \includegraphics[width=0.20\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
\vspace{-20pt}
\end{wrapfigure}
an upper bound on the lines that they can transfer. This can be useful to utilize available
parallelism after encountering a command that cannot be parallelized
(\eg \ttt{sha1sum}). The schematic on the right illustrates this scenario.

@@

Finally, the edges that don't refer to a resource and do not start
from a node in the graph represent the graph input, while the edges
that don't refer to a resource and do not point to a node in the graph
are its outputs.

>>>>

% \begin{lstlisting}[language=sh,float=h,numbers=none]
% cat doc.ms |                   # print input file
% groff -t -e -mandoc -Tascii |  # remove formatting
% col -bx |                      # remove backspaces 
% tr A-Z a-z |                   # convert to lower case
% tr -d '[:punct:]' |            # remove punctuation
% sort |                         # put words in alphabetical order
% unique |                       # remove duplicate words
% comm -13 /usr/share/dict -     # report words not in dictionary 
% \end{lstlisting}
% \smallskip
\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$[p_1]$$)
  sort | uniq -c | sort -rn | head 5 > out
\end{lstlisting}
% \smallskip
% \noindent
Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
Combining several features, \unix makes small tasks easy to express;
  for a user with \emph{one} computer and small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
% composition allows the composition of general primitives

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
%\begin{wrapfigure}{r}{0.06\textwidth}
%  \vspace{-12pt}
%  % \begin{center}
%    % \includegraphics[width=0.48\textwidth]{gull}
%    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
%  % \end{center}
%  \vspace{-30pt}
%\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
% \smallskip

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  ls -l | wc -l                          [($$p_2$$)]
\end{lstlisting}
% \smallskip

Commands can be tuned by two mechanisms, namely com\-mand-line options and environment variables.
Options control the execution of a program---\eg changing the \ttt{sort}ing order.
The shell does not have any visibility into these options; 
  after it expands special characters such as \ttt{~} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.

@@

% At a high level, POSIX defines several file types:
% (-) regular files, storing lines of text or binary data,
% (d) directories, grouping together multiple other files,
% (l) symbolic links, pointing to other files,
% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
% \smallskip

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  ls | sed 's/1/2/' | > out.t 2> err.t    [($$p_3$$)]
\end{lstlisting}
%\smallskip

@@

% \smallskip
\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  cat * | tee >(wc -l > l.t) | zip > a.z   [($$p_4$$)]
\end{lstlisting}
% \smallskip

@@

% \smallskip
\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  read | { calc | square; } >/dev/fd/0     [($$p_5$$)]
\end{lstlisting}
% \smallskip

@@

The core of \sys is a dataflow graph model, inspired by the ones used in popular distributed stream processing systems~\sx{related}.
% such as MapReduce \cite{mapreduce:08} and Apache Spark \cite{spark:12}).
These systems introduce data parallelism by parallelizing nodes of the dataflow graph.
Unfortunately, % their optimizations in many of these systems 
 they often do not preserve the model's semantics, blurring the lines between specification, optimization, and implementation.
In contrast, \sys's dataflow model inherently supports the data parallelism found in shell pipelines.
More specifically, it develops a set of parallelization-exposing optimizations represented as semantics-preserving graph transformations, effectively exposing data parallelism as task parallelism.
The \sys optimizer then applies these transformations in an iterative way, optimizing the program to exploit the available computational resources.

@@

%% A very important consequence of this is that we can iteratively apply
%% optimizations that expose data parallelism, or even revert them when
%% the resulting dataflow graph is too wide. This flexibility unifies the
%% unoptimized and optimized versions of a computation, enabling thinking
%% about them with the same conceptual tools.

In order to effectively apply optimizations and distribute a shell script, \sys first translates it to a manipulable representation.
\sys first introduces a variant of the standard dataflow graph model as a candidate intermediate representation~\sx{graph-components}.
The main benefit of this model is that it exposes both task and data parallelism.
It then defines a set of optimizations that expose and exploit the data parallelism of commands in \sta and \pur as semantics-preserving graph transformations~\sx{ir:transformations}.

@@

As mentioned earlier~\sx{bg:pipelines}, the two main abstractions of the
shell are (i) files (or pipes, \emph{viz.} correspondence between files and pipes in \S\ref{bg:pipelines}) that contain data, and (ii) commands that communicate
through these files. Based on that, \sys's graph model defines nodes to represent commands and edges to represent files.

\paragraph{Nodes---Commands}

For a set $D$, we write $D\kstar$ to denote the set of all finite
words over $D$. For words $x, y \in D\kstar$, we write $x \cdot y$
or $xy$ to denote their concatenation. We write $\eps$ for the empty
word. We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$. It is easy to see that $\leq$ is a partial order, often called the \emph{prefix order}.

@@

\noindent
This command takes a stream of lines from its standard input and
returns a stream of lines on its standard output.

An important
observation is that commands might read their input
stream from a sequence of different sources. An illustrative example
is the command \ttt{cat}, which reads its input from a combination
of files and its standard input:

@@

\noindent
In the example above, \ttt{cat} reads file \ttt{x} until it encounters the EOF
character, then reads from \ttt{stdin} until in encounters EOF, and finally reads from
\ttt{y}.

In order to handle this particularity, nodes in the
dataflow graph can have many incoming edges, that are \emph{ordered} in a
sequence. This is different from standard dataflow
graph representations where the semantics of multiple incoming edges
is either zipping the inputs into pairs, or interleaving them arbitrarily.

@@

Nodes may read from more files than just the files of their
input stream (resp. write to more files than their output stream).
However, these files are static (\eg used for configuration or
logging) and do not represent streams of data, and therefore are not
considered part of the dataflow graph.
For now, we assume that these static
files can only be accessed by one node, thus ensuring that they do not
interfere with the distributed implementation.
%% Because of the assumption in section Front End \ref{}, that each file
%% exists only once in the dataflow graph, we can safely assume that
%% reading and writing to these static files does not interfere with the
%% distributed implementation, and doesn't alter the behaviour of the
%% program.

\paragraph{Edges---Files}

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D\kstar$. They could either refer to a specific named file in the file
system, or just be FIFO pipes used for interprocess communication
%% \TODO{Are there any other file types, such as URLs, ...?}.

@@

In order to expose data parallelism from the output of a command, files that
do not point to a resource can have
\begin{wrapfigure}[3]{r}{0.20\columnwidth}
\vspace{-7pt}
  \includegraphics[width=0.20\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
\vspace{-20pt}
\end{wrapfigure}
an upper bound on the lines that they can transfer. This helps utilize available
parallelism after encountering a command that cannot be parallelized
(\eg \ttt{sha1sum}). The schematic on the right illustrates this scenario.

@@

Finally, edges that do not refer to a resource and do not start
from a node in the graph represent the graph input;
edges that do not refer to a resource and do not point to a node in the graph are its outputs.
